FastSRM: A fast, memory efficient and identifiable implementation of the shared response model

### 2022-08-04 paper/main.tex
pass by BT

\begin{abstract}
  The shared response model (SRM) provides a simple but effective framework to analyze
  fMRI data of subjects exposed to naturalistic stimuli.
  %
  However when the number
  of subjects or runs is large,
  % \bt{True, but we the experiments don't make it so explicit ?}
  fitting the model requires a large amount of memory
and computational power, which limits its use in practice.
%
Furthermore, SRM is
not identifiable, which makes the shared response difficult to interpret.

In this work, we implement an identifiable version of SRM and show on real data
that it improves the stability of the recovered shared response.
%
We then introduce FastSRM, which relies on a dimension reduction step that we
prove to yield the same solution as the original algorithm.
%
We show experimentally using synthetic and real fMRI data
that FastSRM is considerably faster and more memory efficient
than current implementations.

@@

Another approach is to learn jointly a set of reference time courses  and the spatial maps in
an unsupervised way.
%
These reference time courses formally replace the design matrix in the
spatio-temporal decomposition of the data, but cannot be considered as
a formal design matrix.
%
In the case where the dataset contains only one subject,
independent component analysis~\cite{jutten1991blind} (ICA) is the method of choice.
%
ICA assumes that components are independent, a defendible assumption
which ensures identifiability up to permutation and scaling of the
model and can be fitted efficiently.
%
When the dataset contains multiple subjects undergoing the same
protocol, a natural assumption is to assume that the temporal reference is
shared across subjects.
%
Many different methods can produce shared components from different
subjects.
%
Some assume independent
components~\cite{richard2021model}~\cite{richard2020modeling}~\cite{varoquaux2009canica}~\cite{calhoun2001method},
while others are only based on second order information such as
Multiset Canonical Correlation Analysis~\cite{via2011joint} or the
Shared Response Model (SRM)~\cite{chen2015reduced}.
%

@@

What makes the analysis of fMRI data particularly difficult is that
the number of features (voxels) is usually much
larger than the number of available samples (time frames).
Yet, SRM has initially been designed to work within regions of interest using
only few subjects.
%
When using full brain data, computational costs become
high.
%
In addition, memory requirements are difficult to meet since the full dataset
needs to hold in memory.
%
Lastly, another issue of SRM is its non-identifiability (see the proof
in~\cite{richard2020modeling} Appendix D also reported in Appendix~\ref{app:unidentifiability} for completness) which makes it challenging to
interpret the extracted shared components as they are defined up to a rotation.
%

@@

\begin{figure}
  \centering
  \includegraphics[scale=0.3]{figures/srm/conceptual_figure33.png}
  \caption{\textbf{Shared response model}: The raw fMRI data are modeled as a
    weighted combination of subject-specific spatial components with additive
    noise.
    %
    The weights are shared between subjects and constitute the shared response
    to the stimuli.
    %
  }
  \label{fig:srm:conceptual_figure}
\end{figure}

@@

The complexity of Deterministic SRM is in $\bigO(\mathrm{n_{iter}} mpvn)$, where
 $\mathrm{n_{iter}}$ is the number of iterations.
%
We monitor the convergence by computing the $\ell_{\infty}$ norm of the
gradient.
%
Note that we can monitor the gradient without any increase in complexity.
%
Indeed, after the updates with respect to each mixing matrix have been
carried out, the gradient with respect to $(W_i)_{i=1}^m$ is zero and therefore
to compute the $\ell_{\infty}$ norm of the gradient we only need the gradient
with respect to $S$: $m S - \sum_i A_i^{\top}\xb_i$ where the right hand side is
used in the updates of $S$.
%
The algorithm is stopped when the
gradient falls below a chosen tolerance.
%

@@

After preprocessing, the reduced representation $\zb_i$ is used instead of the
original data $\xb_i$ yielding a time complexity of $\bigO(
\mathrm{T_{preprocessing}} + \mathrm{n_{iter}} mpnr)$.
%
Let us highlight that an experiment is often run multiple times,
typically when the analysis requires cross-validation procedures.
%
In these cases,
the pre-processing is performed only once and the apparent complexity becomes
$\bigO(\mathrm{n_{iter}} mpnr)$ which is faster than SRM by
a factor of $\frac{v}{r}$.
%
The number of components in large spatial decompositions is about $r=1000$ and in full brain data,
the number of voxels is about $300~000$ so that $\frac{v}{r}$ is typically about
$300$.
%
It remains to be shown how to draw a correspondence between FastSRM
and SRM, which we do in the following section.
%

@@

We empirically see that the optimal approach is equivalent to using no spatial decompositions in
terms of performance.
%
This is predicted by the theory in section~\ref{sec:optimal_atlas},
where we demonstrate that these two algorithms yield exactly the same
output from a given input and initialization.
%
% %
% Among approximate methods, the projection
% method yields the best results but are quite far away from optimal results.
%
% %
We also see that ProbSRM gives much better results than DetSRM.
%
In terms of fitting time, FastSRM is about a thousand time faster than SRM after
100 iterations.
%
When no spatial decomposition is used, the number
of iterations has a very strong impact on performance while it has a small impact
when an optimal spatial decomposition is used.
%
Lastly, looking at the convergence curves, we see that even after $100$ iterations, algorithms did
not fully converge.
%
This means that in practice a larger number of
iterations is needed, which would yield an even higher difference in fitting time
between methods using no spatial decomposition and methods using an optimal
spatial decomposition.
%

@@

\section{Implementation and reproducibility}
Our work is fully reproducible and our code is tested and well documented. All
material is available in our github repository \url{https://github.com/hugorichard/FastSRM}.

@@

While brainiak and nilearn are useful to reproduce some of our experiments, they
are not strong dependencies and FastSRM only needs Numpy ($\geq 1.12$), Scipy
($\geq 0.18.0$), Matplotlib ($\geq 2.0.0$), Scikit-learn ($\geq 0.23$), Pytest
($\geq 6.2.5$) , and Joblib ($\geq 1.1.0$).

@@

FastSRM inherits from all the weaknesses of SRM including the fact that
mixing matrices are assumed to be orthogonal, which is debattable in practice.
Another desirable property of such decompositions is the spatial smoothness of the components. While it is often observed that SRM components are smooth, nothing imposes such a constraint in the model. We highlight that the optimal spatial decomposition is only optimal with respect to the SRM model. More precisely, it gives a reduced representation such that applying the SRM model on the reduced representation or on the full data gives the same result. Whether this representation is interpretable or not is irrelevant to us, since it is only a means to accelerate the SRM algorithm. However, as this representation is not unique, it may be interesting to find the “most interpretable” optimal spatial decomposition. Note that this would not change the end result of the SRM algorithm but could be of interest for interpretation purposes.
%
It is an open question whether dimension reduction and learning of the shared
components could be done jointly and efficiently without assuming orthogonal
mixing matrices.
%
Our approach based on optimal spatial decomposition shows that SRM-like methods do not need the
full data to provide accurate estimates of shared components. We believe that such
insights may guide the design of future methods.

>>>>

\begin{abstract}
  The shared response model (SRM) provides a simple but effective framework to analyze
  fMRI data of subjects exposed to naturalistic stimuli.
  %
  However when the number
  of subjects or runs is large,
  fitting the model requires a large amount of memory
and computational power, which limits its use in practice.
%
Furthermore, SRM is
not identifiable, which makes the shared response difficult to interpret.

In this work, we implement an identifiable version of SRM and show on real data
that it improves the stability of the recovered shared response.
%
We then introduce FastSRM, that relies on a dimension reduction step and yields the same solution as the original algorithm.
%
We show experimentally using synthetic and real fMRI data
that FastSRM is considerably faster and more memory efficient
than current implementations.

@@

Another approach is to learn jointly a set of reference time courses  and the spatial maps in
an unsupervised way.
%
These reference time courses formally replace the design matrix in the
spatio-temporal decomposition of the data, but cannot be considered as
a design matrix, as it is data-driven.
%
In the case where the dataset only contains  one subject,
independent component analysis~\cite{jutten1991blind} (ICA) is the method of choice.
%
ICA assumes that components are independent, a defendible assumption
which ensures identifiability up to permutation and scaling of the
model and can be fitted efficiently.
%
When the dataset contains multiple subjects undergoing the same
protocol, a natural assumption is to assume that the temporal reference is
shared across subjects.
%
Many different methods can produce shared components from different
subjects.
%
Some assume independent
components~\cite{richard2021model}~\cite{richard2020modeling}~\cite{varoquaux2009canica}~\cite{calhoun2001method},
while others are only based on second order information such as
Multiset Canonical Correlation Analysis~\cite{via2011joint} or the
Shared Response Model (SRM)~\cite{chen2015reduced}.
%

@@

What makes the analysis of fMRI data particularly difficult is that
the number of features (voxels) is usually much
larger than the number of available samples (time frames).
Yet, SRM has initially been designed to work within regions of interest using
only few subjects.
%
When using full brain data, computational costs become
high.
%
In addition, memory requirements are difficult to meet since the full dataset
needs to hold in memory.
%
Lastly, another issue of SRM is its non-identifiability (see the proof
in~\cite{richard2020modeling} Appendix D also reported in Appendix~\ref{app:unidentifiability} for completeness) which makes it challenging to
interpret the extracted shared components as they are defined up to a rotation.
%

@@

\begin{figure}
  \centering
  \includegraphics[scale=0.4]{figures/srm/conceptual_figure33.png}
  \caption{\textbf{Shared response model}: The raw fMRI data are modeled as a
    weighted combination of subject-specific spatial components with additive
    noise.
    %
    The weights are shared between subjects and constitute the shared response
    to the stimuli.
    %
  }
  \label{fig:srm:conceptual_figure}
\end{figure}

@@

The complexity of Deterministic SRM is in $\bigO(\mathrm{n_{iter}} mpvn)$, where
 $\mathrm{n_{iter}}$ is the number of iterations.
%
We monitor the convergence by computing the $\ell_{\infty}$ norm of the
gradient.
%
Note that we can monitor the gradient without any increase in complexity.
%
Indeed, after the updates with respect to each mixing matrix have been
carried out, the gradient with respect to $(W_i)_{i=1}^m$ \textbf{XXX $(W_i)_{i=1}^m$ not introduced so far} is zero and therefore
to compute the $\ell_{\infty}$ norm of the gradient we only need the gradient
with respect to $S$: $m S - \sum_i A_i^{\top}\xb_i$ where the right hand side is
used in the updates of $S$.
%
The algorithm is stopped when the
gradient falls below a chosen tolerance.
%

@@

After preprocessing, the reduced representation $\zb_i$ is used instead of the
original data $\xb_i$ yielding a time complexity of $\bigO(
\mathrm{T_{preprocessing}} + \mathrm{n_{iter}} mpnr)$.
%
Let us highlight that an experiment is often run multiple times,
typically when the analysis requires cross-validation procedures.
%
In these cases,
the pre-processing is performed only once and the apparent complexity becomes
$\bigO(\mathrm{n_{iter}} mpnr)$ which is faster than SRM by
a factor of $\frac{v}{r}$.
%
The number of components in large spatial decompositions is about $r=1000$ and in full brain data,
the number of voxels is typically in the  $50~000 - 400~000$ range, so that $\frac{v}{r}$ is typically about
$50$ to $400$.
%
It remains to be shown how to draw a correspondence between FastSRM
and SRM, which we do in the following section.
%

@@

We empirically see that the optimal approach is equivalent to using no spatial decomposition in
terms of performance.
%
This is predicted by the theory in section~\ref{sec:optimal_atlas},
where we demonstrate that these two algorithms yield exactly the same
output from a given input and initialization.
%
% %
% Among approximate methods, the projection
% method yields the best results but are quite far away from optimal results.
%
% %
We also see that ProbSRM gives much better results than DetSRM.
%
In terms of fitting time, FastSRM is about a thousand time faster than SRM after
100 iterations.
%
When no spatial decomposition is used, the number
of iterations has a very strong impact on performance while it has a small impact
when an optimal spatial decomposition is used.
%
Lastly, looking at the convergence curves, we see that even after $100$ iterations, algorithms did
not fully converge.
%
This means that in practice a larger number of
iterations is needed, which would yield an even higher difference in fitting time
between methods using no spatial decomposition and methods using an optimal
spatial decomposition.
%

@@

\section{Implementation and reproducibility}
Our work is fully reproducible and our code is tested and well documented. All
material is available in the \url{https://github.com/hugorichard/FastSRM} Github repository.

@@

While Brainiak and Nilearn are useful to reproduce some of our experiments, they
are not strong dependencies and FastSRM only needs Numpy ($\geq 1.12$), Scipy
($\geq 0.18.0$), Matplotlib ($\geq 2.0.0$), Scikit-learn ($\geq 0.23$), Pytest
($\geq 6.2.5$) , and Joblib ($\geq 1.1.0$).

@@

FastSRM inherits from all the weaknesses of SRM including the fact that
mixing matrices are assumed to be orthogonal, which is debattable in practice.
Another desirable property of such decompositions is the spatial smoothness of the components. While it is often observed that SRM components are smooth, nothing imposes such a constraint in the model.
%
We highlight that the optimal spatial decomposition is only optimal with respect to the SRM model.
%
More precisely, it gives a reduced representation such that applying the SRM model on the reduced representation or on the full data gives the same result.
%
Whether this representation is interpretable or not is irrelevant here, since we only focus on accelerating the SRM algorithm.
%
However, as this representation is not unique, it may be interesting to find the “most interpretable” optimal spatial decomposition. Note that this would not change the end result of the SRM algorithm but could be of interest for interpretation purposes.
%
It is an open question whether dimension reduction and learning of the shared
components could be done jointly and efficiently without assuming orthogonal
mixing matrices.
%
Our approach based on optimal spatial decomposition shows that SRM-like methods do not need the
full data to provide accurate estimates of shared components. We believe that such
insights may guide the design of future methods.
